---
title: "Data 622 Homework 3: Palmer Penguins, Loan Approval Data EDA Analysis"
author: "Group 6: Alexander Ng, Scott Reed, Philip Tanofsky, Randall Thompson"
date: "Submitted by 04/09/2021"
output:
  html_document:
    df_print: paged
    highlight: pygments
    number_sections: yes
    theme: readable
    toc: yes
    toc_depth: 2
    toc_float: no
    fontsize: 12
  pdf_document:
    toc: yes
    toc_depth: '3'
    number_sections: true
    df_print: kable
    highlight: tango
editor_options:
  chunk_output_type: inline
fontsize: 11pt
urlcolor: blue
---

```{r setup, include=FALSE}
# Your libraries go here

library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)

library(caret)    # Model Framework
library(skimr)    # Used for EDA
library(klaR)     # Implemented KNN and Naive Bayes models, etc
library(class)    # used for KNN classifier

# PLEASE ADD YOUR R LIBRARIES BELOW
# ------------------------------
library(tidymodels)
library(VSURF) 
library(dplyr)
library(mice)

library(vip)


# ---------------------------------
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

```

# Introduction {-}

This document discusses analyses of two datasets, the Palmer Penguin dataset and a Loan Approvals dataset prepared by Group 6.  We divide the document into five parts and adopted two key principles to undertaking this analysis:  

First, group has developed a system of checks and balances in preparing each model's output.  A primary and secondary author cross validate each model (pun intended) by  independently coding and reconciling outputs.  Two co-authors also bring a wider perspective on discussion and diagnostics.   Afterwards, the primary author drafts the text for inclusion in this document.  

Second, because four of the five sections deal with the Loan Approvals data, we decided to adopt a common dataset for all model analyses.  Data wrangling and preparation are done once only. Consistency of the data in the last four parts is essential to assess model performance.

**Section 1** analyzes the Palmer Penguin dataset using the KNN model to predict `species`.  The authors are Alexander Ng (primary) and Randy Thompson (secondary).  

**Section 2** conducts an exploratory data analysis and defining the common data set for modeling.  The authors are Randy Thompson (primary) with contributions by all other members.

**Section 3** analyzes the Loan Approvals by Decision Tree.   The authors are Randy Thompson (primary) and Philip Tanofsky (secondary).  

**Section 4** analyzes the same Loan Approvals with a Random Forest model.  The authors are Scott Reed (primary) and Alexander Ng (secondary). 

**Section 5** analyzes the same Loan Approvals with a Gradient Boosting approach.  The authors are Philip Tanofsky (primary) and Scott Reed (secondary).  

**Section 6** concludes with a discussion of the model performance and an appraisal of the merits of each result.  We also consider the role of model driven prediction in a loan approvals business context.

**Section 7** presents our `R` code and technical appendices and references.

# Penguins and the KNN algorithm

Following the data cleaning approach taken in prior assignments, we exclude observations where the feature `sex` is undefined.
We also exclude the two variables `island` and `year` which do not seem helpful in prediction analysis.  We follow the variable selection process described in a prior assignment [here](https://rpubs.com/Fixed_Point/728197).

This still gives a substantial subset $333$ out of $344$ original observations.
The below table illustrates sample rows of the cleansed dataset used in the subsequent KNN analysis.


```{r cleaned-subset}
library(palmerpenguins)

# The final dataset excludes observations with missing sex and drops the island and year variables.
pc = penguins %>% filter( is.na(sex) == FALSE) %>% dplyr::select( -one_of("island", "year") )

head(pc) %>% kable(caption="Scrubbed penguin data set") %>% kable_styling(bootstrap_options = c("striped", "hover"))

```

We obtain `r nrow(pc)` rows of observations with 6 columns of explanatory and response variables.

## Data Transformation for KNN

The next step is to normalize the data so that each quantitative variables have a mean zero and standard deviation equal to 1.
However, we need to also include the `sex` variable in the K-nearest neighbor model.   Normalization ensures that each variable has equal influence on the nearest neighbor calculation.   Otherwise, a quantitative variable whose scale is much larger than the other variables would dominate the Euclidean distance calculation used in the KNN model.

However, we also need to include the `sex` variable as a feature in the KNN model.  Categorical variables pose a challenge in the KNN context because they need to be mapped to a numerical equivalent.  In the case of this data set, two facts are key:

a) the original data set gathered penguins by breeding pairs - thus the number of males and females is very balanced.
b) the variable is binary.  Thus, two equal and opposite values can accomplish the goal of normalization.

For these reasons, we do not normalize `sex` but use the mapping function $f$ defined as:

$$\begin{equation}
f(\text{sex}) = 
   \begin{cases}
    1 & \text{sex} = \text{male} \\
    -1 & \text{sex} = \text{female} \\
   \end{cases}
\end{equation}
$$
Because the data set is gender balanced, $E[f(\text{sex})] \approx 0$ and $Stdev(f(\text{sex})) \approx 1$.
Hence, $f(\text{sex})$ is nearly normalized.  With these transformations, the KNN dataset is prepared for model evaluation.
We illustrate the top 5 rows of the training data set below.  Note that the sampled indices are displayed in the left most column as row values in the resulting dataframe.

```{r knn-split-data}

set.seed(10)

#  Construct the standardized dataset with the response variable, the 4 quantitative variables normalized 
#  with the scale function and the sex variable normalized with the 1,-1 assignment explained previously.
#
standardized = cbind( pc[,1], scale( pc[, 2:5] ) , sex = data.table::fifelse(pc$sex == 'male' , 1 , -1) )

# Define an 80-20 split of the training and test data.
# -------------------------------------------------------------------------------------
training.individuals = createDataPartition(standardized$species, p= 0.8 , list = FALSE)

# X variables include bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g,  sex (converted to -1,1)
# Y variable is just the species class
# train data is the standardized subset of the full data set
# test data is the complement of the training data set.
# -----------------------------------------------------------------
train.X  =  standardized[ training.individuals,  2:6]
test.X   =  standardized[-training.individuals,  2:6]
train.Y  =  standardized[ training.individuals,    1]
test.Y   =  standardized[-training.individuals,    1]

head(train.X ) %>% kable(caption="Normalized Training Data - 1st Few Rows", digits = 2 ) %>% 
  kable_styling(bootstrap_options = c("striped", "hover") )
```


## Cross Validation and Parameter Tuning with KNN

We will use `caret` to test a range of possible values of $k$ and select the optimal $k$ based on minimizing the cross validation error.   We did the pre-processing previously using `scale` so there is no need to invoke further data transformation inside `caret` pre-processing function calls.

`trainControl` allows us to define the training parameters.

`tuneLength` decides the range of allowed $k$ value to compute against the model.

We choose a 5-fold cross validation method to ensure the test set is sufficiently large.

```{r knn-model-plot}
model = train( species ~. , data = standardized , method = "knn" ,
         trControl = trainControl("cv", number = 5 ), tuneLength = 10 )

# Plot the model output vs. varying k values.
plot(model)
```

Using the above model plot which shows the average cross validation accuracy at each parameter $k$ used in the KNN model, we see that the best kappa and accuracy occur at $k=5$ nearest neighbors.  This is evident when we display the (cross validation) model output below.

```{r knn-model-result}
model

# The best tuning parameter k that 
# optimized model accuracy is:
#model$bestTune
```

## KNN Model Output and Performance

To assess model performance and investigate the suitability of our choice of $k=5$, we consider two approaches.
First we examine the statistical output of the confusion matrix.   The accuracy and kappa ought to look attractive in the model selection.   Second, we examine the smoothness of the decision boundary by plotting the decision boundary for $k=1$ and $k=5$.
If the optimal $k$ shows a relatively smooth boundary, this is supporting evidence that the model parameters are reasonable.

First, we construct the confusion matrix of one specific implementation of the KNN model for the train-test split following the same proportions as the cross-validation study.  We specifically use the `caret` interface rather than the simpler `class` interface to ensure consistency across multiple models in future.


```{r knn-predmulti}

# Simpler interface using the knn function within class.
knn.pred5 = knn(train.X, test.X, train.Y , k = 5, prob = FALSE) 
cm5 = confusionMatrix(knn.pred5, test.Y) 

```

```{r knn-train}

# This model approach uses tuneGrid with a single value to calculate kNN 
# only at a single number of nearest neighbors.
# ------------------------------------------------------------------------------
x2 = train( species ~ . , data = standardized[training.individuals,] , 
            method = "knn", 
            trControl = trainControl(method="none"), 
            tuneGrid=data.frame(k=5) )

pred2 = predict(x2, newdata=test.X)

(cmknn2=confusionMatrix(pred2, test.Y))
```


We see in the confusion matrix output above that accuracy is 98.46% and kappa is 97.6%.  These are consistent with the cross validation results.   We conclude that KNN is a very effective classifier for penguin gender.  Only 1 penguin in 65 was misclassified:  a Chinstrap was predicted to be an Adelie penguin.


```{r build-decision-grid}

# We visualize a projection of the decision boundary flattened to 2 dimensions to illustrate
# the smoothness of the KNN boundary at different values of k.
# First, choose the variables, then build a sampling grid.
# ------------------------------------------------------------------------
pl = seq(min(standardized$bill_length_mm), max(standardized$bill_length_mm), by = 0.05)
pw = seq(min(standardized$bill_depth_mm), max(standardized$bill_depth_mm), by = 0.05 )

# The sampling grid needs explicit values to be plugged in for the variables
# which are not the two projected variables.  Since variables are standardized to
# mean 0 and variance 1, we choose zero for flipper_length and body_mass because they are continuous 
# and the mean and mode are close.
# However, for sex, we choose male because the average sex = 0 is not the mode.
# In practice, sex = 0 gives weird decision boundaries in the projected plot.
# ------------------------------------------------------------------------------
lgrid = expand.grid(bill_length_mm=pl , bill_depth_mm = pw , flipper_length_mm = 0, body_mass_g = 0 , sex = 1 )

knnPredGrid = predict(x2, newdata=lgrid )

num_knnPredGrid = as.numeric(knnPredGrid)

num_pred2 = as.numeric(pred2)

test = cbind(test.X, test.Y)
test$Pred = num_pred2
```

Next, we visualize the decision boundary of our KNN model to check for model suitability.  Custom code needs to be written as we are not aware if this plot type is currently provided by any R package.

Adapting the R recipe for KNN decision boundary plots found in the stackoverflow page [here](https://stackoverflow.com/questions/32449280/how-to-create-a-decision-boundary-graph-for-knn-models-in-the-caret-package)
we see below that the decision boundary for $k=5$ nearest neighbor algorithm is reasonably smooth and not overfitted to the data.

```{r knn-5-decision-boundary}
ggplot(data=lgrid) + stat_contour(aes(x=bill_length_mm, y= bill_depth_mm, z = num_knnPredGrid), bins = 2 ) +
  geom_point( aes(x=bill_length_mm, y= bill_depth_mm, color = knnPredGrid),  size = 1, shape = 19, alpha = 0.2) +
  geom_point( data= test , aes(x=bill_length_mm, y=bill_depth_mm, color=pred2 ) , size = 4, alpha = 0.8, shape =24 ) +
  ggtitle("KNN Decision Boundary for Penguins Data with k=5 neighbor")

```

However, the same KNN decision boundary for $k=1$ shows a much more twisted curve.
The border between the Chinstrap and Adelie regions in the upper middle half shows a zig-zag pattern characteristic of overfitting.  So we reject the choice of $k=1$ as the model even though both  $k=5$ and $k=1$ KNN models have **identical** confusion matrices.  ( The $k=1$ confusion matrix is omitted for brevity. )  This means model selection requires considering factors not captured by the confusion matrix.




```{r knn-1-decision-boundary}

# This model approach uses tuneGrid with a single value to calculate kNN 
# only at a single number of nearest neighbors.
# ------------------------------------------------------------------------------
x3 = train( species ~ . , data = standardized[training.individuals,] , 
            method = "knn", 
            trControl = trainControl(method="none"), 
            tuneGrid=data.frame(k=1) )  # Now we use a k=1 nearest neighbor algorithm which is overfitted.

pred3 = predict(x3, newdata=test.X)


knnPredGrid3 = predict(x3, newdata=lgrid )

num_knnPredGrid3 = as.numeric(knnPredGrid3)

num_pred3 = as.numeric(pred3)

test = cbind(test.X, test.Y)
test$Pred = num_pred3

ggplot(data=lgrid) + stat_contour(aes(x=bill_length_mm, y= bill_depth_mm, z = num_knnPredGrid3), bins = 2 ) +
  geom_point( aes(x=bill_length_mm, y= bill_depth_mm, color = knnPredGrid3),  size = 1, shape = 19, alpha = 0.2) +
  geom_point( data= test , aes(x=bill_length_mm, y=bill_depth_mm, color=pred3 ) , size = 4, alpha = 0.8, shape =24 ) +
  ggtitle("KNN Decision Boundary for Penguins Data with k=1 neighbor")
```



# Loan Approvals:  Exploratory Data Analysis


##  Assessing the Raw Data

Our EDA shows the raw loan approvals dataset has varying level of quality.

Complete variables include:  `Loan_ID`,  `Education`, `Property_Area`, `Loan_Status`, `ApplicantIncome`,  `CoapplicantIncome`
Mostly complete variables include:  `Married`, `Dependents`, `Loan_Amount_Term`, `Gender`

Variables with significant gaps include:   `Self_Employed`,  `LoanAmount`, `Credit_History`
Of these variables, `Credit_History` is the most problematic and it influences the Loan Approval decision.

$\color{red}{\text{@Randy:  Add your detailed discussion here...}}$


## Defining a Common Dataset for Loan Approvals

In this section, we construct a common loan approval data which will be used for all of the model studies that follow.  There are several steps that we take:

1. It consists solely of the records with no missing values for any relevant column and then exclude useless variables.   We call this subset of observations.  This reduces the number of observations from 614 to 480 which is a 78.2% of the original data set.
While this is a significant reduction in the number of observations the remaining data is still sufficient for fitting models.

$\color{red}{\text{@Scott - can you share the evidence that imputing missing values for Credit History is dicey?}}$
$\color{red}{\text{this argument needs to be supported and justifies our choice of a common data set.}}$

2. Next, we omit one irrelevant column:  the `Loan_ID`.  An identifier which is exogenous to the loan applicant and is assumed to be related solely to computer system processes.

3. Then we add a synthetic variable `Total_Income` which represents the sum of the `ApplicantIncome` and `CoapplicantIncome`.

4. We scale all quantitative variables to mean 0 and variance 1 variables with the `scale` function and retain both the nominal and scaled versions.   The scaled variables are named like their nominal counterparts but prefixed with a lower case `s`.  For example, `sApplicantIncome` has mean 0 and variance 1 whereas `ApplicantIncome` has the monthly nominal income.

5. Further, we convert the character type data columns from *text* to *factor*.  This is importance for `Loan_Status` but also relevant for others like:  `Credit_History`, `Property_Area`, `Gender`, `Married`, `Education`, `Self_Employed`.  In addition, we convert `Dependents` from text to an ordered factor because the number of children is ordered.

Lastly, the common data set is exported to flat file.  Each group member performs their Loan Approval model study by loading the common data set and making subsequent selections or transformations while retaining the same observations.


```{r build-common-dataset, eval=T, fig.width=10}
# This code chunk should only be run to build the common data set.  
# Any model builder should rely on importing the common data set only.
# Thus, set eval=TRUE when generating a new version of the common data set.
# Otherwise, set eval=FALSE to skip this step.
# ------------------------------------------------------------------------

# This code block assumes the raw Loan Approval data file in csv form has been placed in the same folder as this script.
# ----------------------------------------------------------------------------------------
cla <- read_csv("Loan_approval.csv") %>% dplyr::select(-Loan_ID)  # drop the row identifier.

# Add a column for Total Income of Applicant and Co-Applicant
# ---------------------------------------------------------------------------
cla$Total_Income = cla$ApplicantIncome + cla$CoapplicantIncome

mice::md.pattern(cla, rotate.names = T)
```

We see that we have some missing variables including a substantial run on Credit History. As this is likely to be a very important variable, has only two output states, and is not obviously derivable from the features provided, it is not a great candidate for imputation. 


```{r}


# We build a dataset in which all observations have fully populated values.
# ---------------------------------------------------------------------------
cla = na.omit(cla)

# Add mean zero and variance 1 versions of quantitative variables.
# ----------------------------------------------------------
cla %>% mutate( sApplicantIncome = scale(ApplicantIncome), 
                sCoapplicantIncome = scale(CoapplicantIncome),
                sTotal_Income = scale(Total_Income) ,
                sLoanAmount   = scale( LoanAmount)
                ) -> cla

head(cla)

write_csv(cla, "cla.csv") # Write the scaled common loan approvals data set to local disk.  

```


After exporting the common loan approval data, we load it into the R session 
and transform character string data to factors.  This allows some R models to better handle response variables.


```{r load-common-data, eval=T}
cla = read_csv("cla.csv")

# Transform the character data columns into factors
# --------------------------------------------------------
cla$Loan_Amount_Term = factor(cla$Loan_Amount_Term)
cla$Loan_Status = factor(cla$Loan_Status)  # Convert the response to factor
cla$Credit_History = factor(cla$Credit_History)
cla$Property_Area = factor( cla$Property_Area)
cla$Gender = factor( cla$Gender)
cla$Married = factor(cla$Married)
cla$Dependents = ordered(cla$Dependents, levels = c("0" , "1", "2" , "3+") )
cla$Education = factor(cla$Education)
cla$Self_Employed = factor( cla$Self_Employed)

head(cla)
```

When performing modeling of common data set, each member may omit one of the Income variables to avoid multi-collinearity problems involving:  `ApplicantIncome`, `CoapplicantIncome` and `Total_Income` or their scaled equivalents.

$\color{red}{\text{Please remember to drop one or more of the 3 Income variables prior to running your models.}}$


## Conditional Approval Rates by Categorical Variable

$\color{red}{\text{Let's add this section (or its equivalent) to the discussion since
it tells a useful story.}}$

$\color{red}{\text{I wrote some code to do the conditional approval rates.}}$

One simple type of exploratory data analysis is to 
examine the conditional approval rates by values of each categorical variable $C$.  
For example, there are  three subcategories (__Urban__, __Suburban__, __Rural__) of the variable `Property_Area`.  
If we find significant differences in the conditional loan approval rates between __Urban__ and __Rural__ applications,
this suggests that `Property_Area` may help to predict approvals.

The conditional approval rates for each categorical variable is displayed below.

```{r approval-rate, results='asis'}

#  Display the approval rate statistics for each qualitative field
#  in absolute count and percentage terms by category.
# 
display_approval_rate <- function(la , field )
{
    a = la[, "Loan_Status"]
    b = la[, field]
    
    c = table(cbind(a,b))  # Count table
    
    pcts = as.data.frame.matrix(c) /nrow(la) * 100   # Percentage equivalent
    col_sums = colSums(pcts)  # Shows the results by the field

    app_rate = pcts[2,]/col_sums   * 100   # Approval rate assumes row 2 is the "Yes" row.

    yy = rbind( pcts, col_sums, app_rate, colSums(c)) 
    rownames(yy)[1]= paste0( rownames(yy)[1], " (%)")
    rownames(yy)[2]= paste0( rownames(yy)[2], " (%)")
    rownames(yy)[3]= paste0( field, " Column Total %:")
    rownames(yy)[4]="Loan Approval Rate by Column%"
    rownames(yy)[5]="Total Applications:"
    print(yy %>% kable(digits = 1, caption = paste0('Loan Approved vs. ', field , ' by Percent') ) %>%
             kable_styling(bootstrap_options = c("striped", "hover") )  %>%
             row_spec(4, background = "skyblue", bold = TRUE) )
}

cat_vars = c("Property_Area", "Credit_History", "Loan_Amount_Term", "Self_Employed", "Education", "Dependents", "Married", "Gender")

for( field in cat_vars)
{
   display_approval_rate( cla , field)  
   cat("\n")
}

```
We observe large differences in approval rates in the following categorical variables:

1.  `Credit_History`:   Those with credit history have 79.3% approval rate vs. 10.0% for those without.

2.  `Property_Area`:   `Rural` approval rates 61.2% are lower than `Semiurban` at 78.0%

3.  `Married`:   Unmarried applicant approval rate 62.1% is lower than for married at 73.0%

4.  `Education`:  `Not Graduate` approval rates 62.9% are lower than `Graduate` at 70.8%

5.  `Gender`:   Female applicant approval rate 62.8% is lower than for males at 70.6%.

These differences should suggest the most significant categorical variables for prediction purposes.



# Loan Approvals: Decision Tree Models


$\color{red}{\text{Please set.seed() in your section to ensure your R code works as expected. }}$


$\color{red}{\text{Suggest a 80-20 split for training and test data to allow cross model comparison. }}$

## Subsection A

## Subsection B

## Subsection C




# Loan Approvals: Random Forest Models


## Feature Selection

We first run an all variables (save the index variable Loan_ID) in a random forest to use it as a variable selection sieve. We use VSURF as it was the best performing in Speiser (2019). 

```{r, echo=FALSE,results = FALSE, message=FALSE }
set.seed(10)
data_split <- initial_split(cla, prop = .8)
train_data <- training(data_split)
test_data  <- testing(data_split)
use_data <- train_data %>%dplyr::select( -ApplicantIncome, -CoapplicantIncome, -Total_Income, -sApplicantIncome, -sCoapplicantIncome, -LoanAmount)
surfedModel <- invisible(VSURF::VSURF(use_data%>%dplyr::select(-Loan_Status), use_data$Loan_Status, parallel=F))

```
```{r, echo=F}
print(paste("VSURF Final Stage ", colnames(use_data%>%dplyr::select(-Loan_Status))[surfedModel$varselect.pred]))
print(paste("VSURF 2nd Stage " , paste(colnames(use_data%>%dplyr::select(-Loan_Status))[surfedModel$varselect.interp], collapse=", ")))
print(paste("VSURF 1st Stage " , paste(colnames(use_data%>%dplyr::select(-Loan_Status))[surfedModel$varselect.thres], collapse = ", ")))
```
VSURF selects a single variable "model", and does so on both the final and penultimate stage. This suggests that a very parsimonious model indeed might be sufficient. 


```{r fig.cap="'Model' Using only Credit History" }
loans_testing <- na.omit(testing(data_split))
loans_testing["factorCreditHistory"] <- as.factor(as.logical((loans_testing)$Credit_History==1))
loans_testing["factorLoanStatus"] <- as.factor((loans_testing)$Loan_Status=="Y")
loans_testing %>% metrics(truth = factorLoanStatus, estimate = factorCreditHistory)
```
Using this as a "model" we get an accuracy of 83.1% and a Kappa of 0.486. Cohen (1960) would suggest that as an agreement unlikely to be caused by chance (which is important as a model that simply flagged every loan as a success would have an accuracy of .737 but a Kappa of course of 0)

### Model building

If we were to ignore VSURF's suggestion we can try to sweep through the stage 1 recommended features (which are reported in order from most promising to least)

```{r}
loan_testing <- na.omit(testing(data_split))
loan_testing["factorCreditHistory"] <- as.factor(as.logical((loan_testing)$Credit_History==1))
loan_testing["factorLoanStatus"] <- as.factor((loan_testing)$Loan_Status=="Y")
actual_result <- loan_testing %>% metrics(truth = factorLoanStatus, estimate = factorCreditHistory)
ourAccuracy <- tibble('Variables' = 1 , 'Kappa' =actual_result$.estimate[2], 'Accuracy' = actual_result$.estimate[1])
ourAccuracy <- ourAccuracy[-c(1),]
#loan_testing <- testing(data_split) #reset with nas added back in
for(var in 1:length(surfedModel$varselect.thres)){
  loans_rf_rec <-  recipe(Loan_Status~., data = train_data)  %>%
    step_rm(everything(),-all_of(colnames(use_data[(surfedModel$varselect.thres)[1:var]])), -Loan_Status)  %>%
    #step_knnimpute(all_predictors()) %>%
    step_dummy(all_nominal(), -all_outcomes()) %>% 
    prep()
  loan_training<-juice(loans_rf_rec)
  loan_ranger <- rand_forest( mode = "classification") %>%   set_engine("ranger") %>%     fit(Loan_Status ~ ., data = loan_training)
  loan_testing <- loans_rf_rec  %>% bake(testing(data_split)) 
  actual_result <- loan_ranger %>% predict(loan_testing) %>%     bind_cols(loan_testing) %>%     metrics(truth = Loan_Status, estimate = .pred_class)
  ourAccuracy <- ourAccuracy %>% add_row('Variables' = var , 'Kappa' =actual_result$.estimate[2], 'Accuracy' = actual_result$.estimate[1])
}
ourAccuracy %>% ggplot(aes(x=Variables,y=Accuracy)) + ggtitle("Accuracy as variables are added") +geom_point() + ylim(0,1)
ourAccuracy %>% ggplot(aes(x=Variables,y=Kappa)) +ggtitle("Kappa as Variables are added") +geom_point() + ylim(0,.6)
ourAccuracy
```

Perhaps unsurprisingly we don't seem to receive any benefit of adding additional variable candidates, indeed the model actually performs somewhat worse for many options.

Still if we use the first stage of VSURF and a wide parameter grid for the tunable parameters mtry and number of trees  we can attempt to tune a model.

```{r, warning=F}
our_data<-use_data[,colnames(use_data[(surfedModel$varselect.thres)])]
our_data$Loan_Status<-use_data$Loan_Status
loans_rf_rec <-  recipe(Loan_Status~., data = our_data)  %>%
  prep()
loans_grid<-grid_regular(trees(), finalize(mtry(), use_data), levels=5)
loan_training<-juice(loans_rf_rec)
loan_ranger <- workflow()%>% add_recipe(loans_rf_rec) %>% add_model(rand_forest(mtry=tune::tune(), trees = tune::tune(), mode = "classification") %>% set_engine("ranger",importance = "impurity"))
folds <- vfold_cv(our_data, v = 5)
class_metrics <- metric_set(accuracy, kap)
tree_res  <- loan_ranger %>% tune_grid(resamples = folds, grid=loans_grid, metrics=class_metrics)
loans_best <-  tree_res %>% select_best("kap")
final_loans_wf <-loan_ranger %>% finalize_workflow(loans_best)
final_loans_fit <-final_loans_wf %>% fit(data = our_data) 
testing_data<- (testing(data_split)) 
our_data<-testing_data[,colnames(use_data[(surfedModel$varselect.thres)])]
our_data$Loan_Status<-testing_data$Loan_Status
loan_testing <- loans_rf_rec %>% bake(our_data)
actual_result<-final_loans_fit %>% predict(loan_testing) %>%     bind_cols(loan_testing) %>%     metrics(truth = Loan_Status, estimate = .pred_class)
kable(actual_result)
tree_res %>%
    collect_metrics() %>% mutate(trees = factor(trees)) %>% ggplot(aes(mtry, mean), color=trees) +
    geom_line(size = 1.5, alpha = 0.6) +
    geom_point(size = 2) +
    facet_wrap(~ .metric, scales = "free", nrow = 2) +
    scale_color_viridis_d(option = "plasma", begin = .9, end=0)
final_loans_fit %>%      pull_workflow_fit() %>%      vip()
loans_best
```
Our full model has an accuracy of .863 and a Kappa of .624. This is an improvement of our prior model, particularly in Kappa.


## Subsection A

$\color{red}{\text{Please set.seed() in your section to ensure your R code works as expected. }}$

$\color{red}{\text{Maybe use a 80-20 split for training and test data to allow cross model comparison. }}$


## Subsection B

## Subsection C

# Loan Approvals:  Gradient Boosting

## Subsection A

$\color{red}{\text{Please set.seed() in your section to ensure your R code works as expected. }}$

$\color{red}{\text{Maybe use a 80-20 split for training and test data to allow cross model comparison. }}$


## Subsection B

## Subsection C

# Loan Approvals:  Model Performance and Assessment

In this section, we compare the performance of the decision tree, random forest and gradient boosting models on the Loan Approvals dataset.  

## Model Performance

## Using Prediction in Loans Approvals

### Recomendations

It would be very interesting to break down the Credit History Feature into components or at the least have a more continuous variable (say points above or below the FICO score cut off). Our models showed it to have an almost overwhelmingly strong impact on the output model, and presumably it is based on its own more continuous data that could be further analyzed. It seems likely, for example, that loan approval of applicants that do not meet the credit history requirements likely are closer to approval than applicants that seem mostly similar and do not receive approval. However, the data provided does not allow us to state that.



# Appendices

## References

$\color{red}{\text{Put your external references here.}}$

Cohen, J. (1960). A coefficient of agreement for nominal scales. Educational and psychological measurement, 20(1), 37-46.

[UC Business Analytics R Programming Guide - Random Forests](https://uc-r.github.io/random_forests)

Speiser, J. L., Miller, M. E., Tooze, J., & Ip, E. (2019). A comparison of random forest variable selection methods for classification prediction modeling. Expert systems with applications, 134, 93-101.

## Code

We summarize all the R code used in this project in this appendix for ease of reading.


```{r ref.label=knitr::all_labels(), echo=T, eval=F}

```


